---
title: "Programming assignment 3"
format: gfm
date: last-modified
date-format: iso
---

##Load libraries and data
```{r}
library(dplyr)
library(ggplot2)
```

##Descriptive statistics
```{r}
vowel_data <- read.csv("data/vowel_data.csv",
                fileEncoding = "UTF-16",
                encoding = "UTF-8",
                sep = ",",
                stringsAsFactors = TRUE)
```

```{r}
vowel_data |> 
  summarise(
    mean_F1 = mean(f1_cent, na.rm = TRUE),
    sd_F1 = sd(f1_cent, na.rm = TRUE),
    mean_F2 = mean(f2_cent, na.rm = TRUE),
    sd_F2 = sd(f2_cent, na.rm = TRUE),
    mean_tl = mean(tl, na.rm = TRUE),
    sd_tl = sd(tl, na.rm = TRUE),
    n=n()
  ) 

```

##Plots
```{r}
ggplot(vowel_data) +
  geom_boxplot(aes(x = vowel, y = tl))

```
```{r}
ggplot(vowel_data) +
  geom_boxplot(aes(x = language, y = tl))

```
```{r}
ggplot(vowel_data) +
  geom_boxplot(aes(x = language, y = tl, fill = vowel))

```

```{r}
ggplot(vowel_data) +
  geom_boxplot(aes(x = language, y = f1_cent, fill = vowel)) +
  labs(y = "F1 centroid (Hz)",
       x = "Language",
       fill = "Vowel",
       title = "F1 centroid by language and vowel")

```
```{r}
ggplot(vowel_data) +
  geom_boxplot(aes(x = language, y = f2_cent, fill = vowel)) +
  labs(y = "F2 centroid (Hz)",
       x = "Language",
       fill = "Vowel",
       title = "F2 centroid by language and vowel")

```
```{r}

ggplot(vowel_data) +
  geom_point(aes(x = f1_cent, y = f2_cent, color = tl)) +
  labs(x = "F1 centroid (Hz)",
       y = "F2 centroid (Hz)",
       color = "Trajectory length (Hz)")

```
##Questions 
1-This part of the script calculates several key points between a starting point (vonset) and an ending point (voffset). It first measures the total distance or duration between the two points (durationV = voffset - vonset), then uses that value to find positions at 20%, 35%, 50%, 65%, and 80% of the way between the start and end.

2-First, the goal  is to get vonset at the initial of vowel, and then get the end point of the vowel, duration of the vowel ,  and finally calculate the midpoint at different percentage. 

3-It's different form the previous pa because for this we had to calculate the f1, f2, and on top of that the vowel centroids. 
